Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-11 programlogic.tex
+

  $\exists_{\entryf \;, \entrytr \;, \entrytw \;,\entryo} \ldotp$  $\ulcorner$ aligned $\vaddr \urcorner \ast$ L4_L1_PointsTo($\rtv,\entryf,\entrytr,\entrytw,\entryo$) $\ast$ addr_L1$(\vaddr,\entryo) \mapsto_{p} \vpage$.

@@

  (l3M52 maddr entry) \$\sim$\ (l3off maddr entry)  $\mapsto_{a}$ {q2} l3e $\ast$ 
  (l2M52 maddr entry0) \$\sim$\ (l2off maddr entry0) $\mapsto_{a}$ {q3} l2e $\ast$
  (l1M52 maddr entry1) \$\sim$\ (l1off maddr entry1) $\mapsto_{a}$ {q4} l1e.

@@

    $\hbox{(\TirNameStyle{{PageLabelAccess}})}$ addr_L1$(\vaddr$,\textsf{l1e}$) \mapsto_{p} \vpage$.

@@

\begin{figure*}
\[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{v,rtv}}\{\textsf{q}\}\vpage \stackrel{\triangle}{:=} 
    \exists \entryo \ldotp
    \sumapaces\rtv\delta \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast 
   \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr
\end{array}
\]
\caption{Virtual-Pointsto for Sharing Pages}
  \label{fig:virtualpointstosharing}
\end{figure*}  
  One possible approach is to abstract away the physical page-table-walk pointsto ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}) from the virtual pointsto relation. To do so, we introduce the protocol separating read-only (fragmental) and authorative (full ownership) accesses to the page tables, specifically $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$. We realize this protocol by embedding it into an already existing \textsf{Iris}'s ghost-map (or view\_maps) construction
  \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  (\mathcal{W}_{64} \; \rightarrow_{\textrm{fin}}\;  (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
 where
  \begin{itemize}
  \item we decorate the fragmental ownership of the ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$) with the physical $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ whose ownership just ensures that the virtual address ($\vaddr$) is mapped to a page (referenced by $\paddr$)
  \item we place the authorative ownership $\mathcal{A}\textsf{PTableWalk}$ of the ghost-map in an address-space invariant ($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}) whose ownership grants sound update on the page table entries
  \end{itemize}

  \begin{figure*}
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace} \stackrel{\triangle}{:=} 
  \exists\;\gammaPred \;\ldotp\; \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \\
  \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr,\rtv,\entryf,\entrytr,\entrytw,\entryo)}
\end{array}
\]
\caption{Global Address-Space Invariant}
  \label{fig:peraspaceinvariant}
  \end{figure*}

\begin{remark}[Exchanging Page-Table-Walk-Summarization Mappings (Tokens) under Mutation of Page Tables ]
  $\mathcal{A}\textsf{PTableWalk}$'s fragmental-ownership represents an individual key-value pair (per-virtual-address) which can be handled as a \textit{look-up} resource for an individual virtual address.
  To build more intuition on the convenience of this protocol, we can think of the use-case \textit{unmapping-pages}. Normally, in the existence of shared page-tables, we need to first collect all the fragmental map pairs, e.g. $\sumwalkabs\vaddr\qfrac\paddr$, then, after making sure of non-existence of \textit{sharing}, we can \textit{unmap} the page. In this case, it would be ok to put $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ and $ \sumwalkabs\vaddr\qfrac\paddr$ side-by-side inside the virtual-pointsto definition in Figure \ref{fig:virtualpointstosharing}. However, this pattern of verifying is not only inconvenient due to requirement of extra bookkeeping, but also violates our soundness argument based on considering  mutation on the shared page-tables -- due to coalescing of page tables etc. --. In other words, physical $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ sitting aside to  $ \sumwalkabs\vaddr\qfrac\paddr$ would be made invalid by any concurrent update on the physical page-table. So there should only be one physical memory resource inside the virtual pointsto definition: from physical page address to the value stored ($   \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr$).
\end{remark}

@@

In Figure \ref{fig:addrswitch}, we see two different address-spaces rooted at R0 and R1 in the pre/post condition. To be put into words, while switching an address-space by executing \lstinline|mov_ctl|) updating the  \lstinline|cr3|, it is expected to mention resources of two different address-spaces in the specification. In the precondition,  \lstinline|cr3| is loaded with R0 and all the virtual-pointsto assertions $a_0 \mapsto_{\textsf{v}}p_0$ and $a_1 \mapsto_{\textsf{v}}p_1$ are in the current-view of the memory and valid in the ambient logic \textsf{vProp}. The other address-space with the root address R1 has its own resources ($b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$) which are invalid under the current address space rooted at R0. This pattern of specification requires resources being mentioned under the \textit{world} they are valid, i.e. a modal-framing within the specification. Once the current address-space becomes the one with root address R1, i.e.  \lstinline|cr3| is loaded with R1, the context-resources $b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$ are loaded into the current-view of the memory, and all the other resources that are valid under R0 are introduced into the modal context rooted at R0
\begin{figure}

\[
  \begin{array}{l}
    \hbox{(\TirNameStyle{ModalAddressSpace)}} \qquad
         [r](P) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{:=} \; \lambda \mathsf{P},\; \mathsf{P r}. 
  \end{array}
  \]
  \label{fig:modaldef}
  \end{figure}

>>>>

  $\exists_{\textsf{l4e l3e l2e l1}} \ldotp$  $\ulcorner$ aligned $\vaddr \urcorner \ast$ L4_L1_PointsTo(rtv maddr l4e l3e l2e l1e) $\ast$ addr_L1$(\vaddr,\textsf{l1e}) \mapsto_{p} \vpage$.

@@

  (l3M52 maddr l4e) \$\sim$\ (l3off maddr l4e)  $\mapsto_{a}$ {q2} l3e $\ast$ 
  (l2M52 maddr l3e) \$\sim$\ (l2off maddr l3e) $\mapsto_{a}$ {q3} l2e $\ast$
  (l1M52 maddr l2e) \$\sim$\ (l1off maddr l2e) $\mapsto_{a}$ {q4} l1e.

@@

    $\hbox{(\TirNameStyle{{PageLabelAccess}})} \qquad$ addr_L1($\vaddr,\textsf{l1e}$) $\mapsto_{p} \vpage$.

@@

\begin{figure*}
\[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{v,rtv}}\{\textsf{q}\}\vpage \stackrel{\triangle}{:=} 
    \exists \paddr\ldotp
    \sumapaces\rtv\delta \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{p} \vpage
\end{array}
\]
\caption{Virtual-Pointsto for Sharing Pages}
  \label{fig:virtualpointstosharing}
\end{figure*}  
  One possible approach is to abstract away the physical page-table-walk pointsto ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}) from the virtual pointsto relation. To do so, we introduce the protocol separating read-only (fragmental) and authorative (full ownership) accesses to the page tables, specifically $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$. We realize this protocol by embedding it into an already existing \textsf{Iris}'s ghost-map (or view\_maps) construction
  \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  (\mathcal{W}_{64} \; \rightarrow_{\textrm{fin}}\;  (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
 where
  \begin{itemize}
  \item we decorate the fragmental ownership of the ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$) with the physical $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ whose ownership just ensures that the virtual address ($\vaddr$) is mapped to a page (referenced by $\paddr$)
  \item we place the authorative ownership $\mathcal{A}\textsf{PTableWalk}$ of the ghost-map in an address-space invariant ($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}) whose ownership grants sound update on the page table entries
  \end{itemize}

  \begin{figure*}
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace} \stackrel{\triangle}{:=} 
  \exists\;\gammaPred \;\ldotp\; \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \\
  \bigast{(\vaddr, \textsf{l1e})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\rtv,\textsf{ maddr, l4e, l3e, l2e, l1e})}
\end{array}
\]
\caption{Global Address-Space Invariant}
  \label{fig:peraspaceinvariant}
  \end{figure*}

\begin{remark}[Exchanging Page-Table-Walk-Summarization Mappings (Tokens) under Mutation of Page Tables ]
  $\mathcal{A}\textsf{PTableWalk}$'s fragmental-ownership represents an individual key-value pair (per-virtual-address) which can be handled as a \textit{look-up} resource for an individual virtual address.
  To build more intuition on the convenience of this protocol, we can think of the use-case \textit{unmapping-pages}. Normally, in the existence of shared page-tables, we need to first collect all the fragmental map pairs, e.g. $\sumwalkabs\vaddr\qfrac\paddr$, then, after making sure of non-existence of \textit{sharing}, we can \textit{unmap} the page. In this case, it would be ok to put $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ and $ \sumwalkabs\vaddr\qfrac\paddr$ side-by-side inside the virtual-pointsto definition in Figure \ref{fig:virtualpointstosharing}. However, this pattern of verifying is not only inconvenient due to requirement of extra bookkeeping, but also violates our soundness argument based on considering  mutation on the shared page-tables -- due to coalescing of page tables etc. --. In other words, physical $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ sitting aside to  $ \sumwalkabs\vaddr\qfrac\paddr$ would be made invalid by any concurrent update on the physical page-table. So there should only be one physical memory resource inside the virtual pointsto definition: from physical page address to the value stored ($ \paddr \mapsto_{p} \vpage$).
\end{remark}

@@

In Figure \ref{fig:addrswitch}, we see two different address-spaces rooted at R0 and R1 in the pre/post condition. To be put into words, while switching an address-space by executing \lstinline|mov_ctl|) updating the  \lstinline|cr3|, it is expected to mention resources of two different address-spaces in the specification. In the precondition,  \lstinline|cr3| is loaded with R0 and all the virtual-pointsto assertions $a_0 \mapsto_{\textsf{v}}p_0$ and $a_1 \mapsto_{\textsf{v}}p_1$ are in the current-view of the memory and valid in the ambient logic \textsf{vProp}. The other address-space with the root address R1 has its own resources ($b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$) which are invalid under the current address space rooted at R0. This pattern of specification requires resources being mentioned under the \textit{world} they are valid, i.e. a modal-framing within the specification. Once the current address-space becomes the one with root address R1, i.e.  \lstinline|cr3| is loaded with R1, the context-resources $b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$ are loaded into the current-view of the memory, and all the other resources that are valid under R0 are introduced into the modal context rooted at R0
\begin{figure}
\[
  \begin{array}{l}
    \hbox{(\TirNameStyle{ModalAddressSpace)}} \qquad
         [r](P) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{:=} \; \lambda \mathsf{P},\; \mathsf{P r}. 
  \end{array}
  \]
  \label{fig:modaldef}
  \end{figure}
